From correlation to causality: statistical approaches to learning regulatory relationships in large-scale biomolecular investigations

### 2015-12-04 manuscript/write_up.tex
incorporated changes from review

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Place any additional macros here.  Please use \newcommand* where
%% possible, and avoid layout-changing macros (which are not used
%% when typesetting).
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
\usepackage{color}
\def\todo#1{{\color{red}[TODO: #1]}}

@@

Causal inference \cite{pearl2009causality} applied to 'omics experiments, elucidates statistical associations (e.g. Pearson correlation, Spearman correlation, mutual information), which result from the underlying biomolecular mechanistic relationships. Modern high-throughput technologies such as mass spectrometry-based proteomics quantify components of the biomolecular systems on a large scale. We will demonstrate the paradoxical result that the large amount of data generated by these experiments make the task of causal inference more difficult.  If only two biologically unrelated analytes were measured simultaneously, there is some probability of a spurious association -- an association between the analytes arising purely from random chance -- though the chance of this happening is low and gets lower as sample size increases. The difficulty of high-throughput data stems from the fact that the more analytes that are measured, the more pairwise associations between analytes are generated, resulting in a much greater chance for the appearance of spurious associations that are just as strong as associations that represent actual biological relationships.  This obfuscates the true signal, leading to increased false discoveries of putative causal events.  This problem can be addressed by refining the biological question, and by improving experimental design in terms of selection of  (1) the subset of analytes, (2) the number of biological replicates, and (3) the type of biological conditions and stresses. Below, we describe in non-technical terms the process of elucidating causal associations from high-throughput data and suggest practical approaches for analysis of large scale datasets.

@@

To describe Step 2 of causal inference, we introduce some terminology. In language of probability and statistics the quantified proteins are called {\it variables}. When two proteins are connected in how they vary, such that knowing the behavior of one protein provides information on the other, they are called {\it dependent}.  If there is no such connection, they are called \{it independent}. {\it Conditional independence} is a special case that is important to causal inference.  Two dependent variables are conditionally independent if, after knowing, or {\it conditioning on} the behavior of third-party variables, they become independent -- knowing one provides no additional information about the other after the information from the third party variables is considered.

If one protein has direct causal influence on another, they are dependent, even after conditioning.  So if we know two proteins are not conditionally independent, we can investigate whether or not they have a causal relation.  However, we do not know the actual dependence relations between proteins, and must use statistical tests for conditional independence to infer them.  These tests take as input the quantifications of statistical association derived in step 1, which in our example is the Spearman correlations as in Figure \ref{mapkInference}B.  Note {\it dependence} is a property of the underlying biology, and {\it statistical association} is a property of the data generated by the biology, they are not the same.

Let's see how this applies to the MAPK signaling cascade. We can examine the nature of statistical association between Raf and Erk.  Figure \ref{mapkInference}C compares Raf to Erk, indicating in circles the biological samples where concentrations of Mek are high (here, set to the top quartile).  {\it Note that when we subset the data to only the samples with high Mek (in statistical language, when we {\it condition} on Mek being high), we can no longer visually detect the association between Raf and Erk}.

Thinking in terms of a statistical test, our null hypothesis is conditional independence between Raf and Erk given Mek.  We take the typical significance threshold of .05 --- p-values must be below this value for us to reject the null.  We discretized the protein values into 3 levels; low, medium, and high.  We then calculate the Pearson chi-squared test statistic -- a value that quantifies the statistical association between Raf and Erk after accounting for Mek's association with both variables (though we use Spearman correlation in the figure, chi-squared is preferred here because using the former in a conditional independence test is non-trivial).  In this case the value of the test statistic is 10.24 with 12 degrees of freedom, which has p-value of .595 -- which doesn't meet our threshold.  Thus we decide in favor of the null hypothesis -- the high p-value tells us we can't be confident the statistical association between Raf and Erk, beyond what is explained by Mek, did not arise spuriously.  Tests for conditional independence between Raf and Mek given Erk, and Mek and Erk given Raf both lead to our rejecting the null (p = .034, p = .027).  Given our decisions, the edge between Raf and Erk is removed, resulting in the middle graph of Figure \ref{mapk}A, a graph of putative conditional dependence relationships.

This result has a mechanistic explanation. As can be seen in Figure \ref{mapk}, the abundance of Erk in each sample is determined by Mek. Therefore, when the abundance of Mek is fixed, Raf does not exert any additional influence on the variability of Erk.

When the experiment quantifies all the key variables in a biological system, Step 2 above elucidates conditional dependence between causally related variables. In the MAPK example, this corresponds to a signaling protein's direct regulators, its direct effectors, and other proteins who share its direct effectors.  However, at this step the direction of the regulation remains unknown. Inference of the direction of the chain of events requires that the experimental design involves external interventions or stresses. Figure \ref{mapkInference}D illustrates the results of Step 3, in the case where an intervention targeted Mek with an inhibitor. The intervention does not affect the concentration of Mek, however it blocks its ability to phosphorylate other proteins.  After this intervention the Raf--Mek relationship is unchanged, while Erk drops to a low level.  From this we can infer that Mek has causal influence on Erk, and since Raf was unaffected by the intervention, that Raf has causal influence on Mek.  With the intervention, we can finally move from the conditional dependence graph in panel A - step 2 to the causal graph in panel A - step 3.

In the general case, computational methods for causal inference follow the workflow  in Figure \ref{mapkInference}A, while scaling it to characterize multiple inter-related variables. Step 1 creates a dense network of pairwise associations. Step 2 reduces this dense network to a sparse network of putative conditional dependencies, using empirical evidence of  conditional independence between pairs of variables.

Finally, Step 3 uses the experimental design, specifically the information regarding the interventions, to evaluate these conditional dependencies as evidence for potential causal events.  See Koller-Friedman \cite{koller2009probabilistic} for a detailed description of these methods and their theoretical underpinnings. Numerous implementations of these algorithms are available, e.g. in the R package {\tt bnlearn} \cite{scutari2009learning}.

@@

Similarly, the increased incidence of spurious correlations impedes the performance of statistical methods in Step 2, which elucidate conditional independences in the data.  The spurious correlations result in more false positives when detecting putative causal conditional dependence relationships. To illustrate, we repeated the previous simulation, again starting with 20 proteins and 100 biological samples, but this time expanding to only 100 proteins.  Instead of finding the highest  spurious correlation between pairs of proteins, we apply the a causal inference-related algorithm described in Margaritis 2003 \cite{margaritis2003learning} (which performs a series of conditional independence tests between the sets of proteins), and count the number of detected conditional dependence relationships.  As before, since we randomly draw protein concentration measurement values from a Gaussian distribution, the values are completely independent, and any conditional dependence relationship reported by this algorithm has no biological justification.  We again repeat these experiments 500 times. Figure \ref{spur_dep} shows the histograms of the counts of false positive detections of conditional dependence.  The results demonstrate that an increase in the number of proteins leads to an increase in false positive detection of putative causal conditional dependence relationships. This means that the computational methods for causal inference will fail for a typical large-scale experiment, because they cannot reliably distinguish the conditional dependence relationships from noise.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=1\textwidth]{figs/spurious_dep.png}}
\caption{As in Figure \ref{spur_corr}, but reporting the counts of false positive detections of conditional dependence.}
\label{spur_dep}
\end{figure}

An additional problem with high-throughput experiments in Step 2 is a relatively small number of biological replicates as compared to the number of analytes. While Step 1 (evaluating the set of pairwise associations) can be carried out with even a small amount of replicates, Step 2 requires that the number of replicates equals or exceeds the number of analytes.  Note the experiments demonstrated in Figure 3 had 100 replicates to its 100 proteins.  While sparse approaches working around this problem are available, quality of their results decreases drastically as the numbers of replicates fall.

Finally, Step 3 of the workflow requires interventions to infer causal relationships from evidence of conditional dependence.  Quantifying a large number of analytes introduces a  challenge for this step as well. As the number of features grows, the number of interventions needed to fully infer causality grows, and eventually performing a sufficient perturbation experiments becomes infeasible.   We showed with the Raf-Mek-Erk example that one intervention was sufficient to infer two causal relationships. Indeed, a set of interventions on $< k$ variables may be sufficient to infer causal events between $k$ analytes. Unfortunately, both the number of biological replicates and the number of interventions im high-throughput profiling experiments is typically small.  Also, if the step 2 results in too many false positives for conditional dependence, this will adversely affect the results of step 3 regardless of the size of the dataset.

@@

\item \textit{Profile more biological replicates.}  The high-throughput measurements should provide more samples from distinct biological sources, which come from a same underlying population, in order to achieve a sufficient statistical power, and distinguish true and spurious associations.  This fact gives advantage to technologies, such as targeted proteomics tools, that quantify fewer analytes but have a higher sample throughput.  Single cell mass cytometry is a technology deserving of special mention, where many thousands of cells per sample provide ample statistical power, though cell samples still need to be drawn from multiple individuals in order to make inferences about the population. See Sachs et al 2005 for an in depth case study of network inference with a single cell dataset. \cite{sachs2005causal}.

\item \textit{Use prior knowledge.} Prior knowledge improves the search for conditional independence and helps to determine causality. The prior knowledge can be in form of known canonical networks, extracted, e.g. from pathway databases such as KEGG. One example of such prior information is the MAPK pathway. The prior information reduces the search space of unknown associations that need to be considered, enables a more effective use of the data, and increases the confidence in newly discovered statistical associations.  Another example of prior knowledge is contextual information, such as spatial or temporal annotations of the quantitative measurements in the cell. The contextual information can be extracted from the literature or from other complementary (and potentially noisy) datasets. The causal inference algorithms can be extended to weigh evidence of conditional dependence, depending on whether the analytes share the same spatial or temporal context.

\item \textit{Select targeted interventions wisely.} Targeted interventions perturb individual components of the biological system.  An example is a small molecule inhibitor, which blocks the causal influence of a specific protein on its downstream components. Although effective, such targeted interventions are limited in number. Therefore, a strategic experimental design would use prior information, prioritize the interventions and the targets, and apply them to parts of the biological system that have most potential for new discovery of regulatory events.  For example, a graph with undirected edges can be inspected, to reveal which nodes have potential to reveal the most causality if perturbed. Such targeted perturbations can be applied iteratively, after an initial statistical analysis revealed areas of the network where causal inference would benefit from extra measurements and data.

\item \textit{Consider broad-scale interventions}. Broad-scale interventions sacrifice specificity of the target to simultaneously perturb many variables in a biological system. One example of broad-scale interventions is varying experimental conditions, in order to activate multiple pathways.  Signals from endocrine, paracrine, and autocrine ligands elicit various signaling responses in hepatocytes, thus interventions that cover this range of signals gives the best picture of the broader causal network of hepatocyte signaling \cite{alexopoulos2010networks}. Similarly, interventions that go beyond receptor-level and perturb multiple components of the system bring cascading causal direct orientation deeper into the network.  Although they do not provide specific information about the downstream effects of stimulation,  broad-scale interventions can provide more causal insight. Therefore, the advantage of this approach is that it may enable elucidation of  causality across the entire system.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Place any additional macros here.  Please use \newcommand* where
%% possible, and avoid layout-changing macros (which are not used
%% when typesetting).
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
\usepackage{color}
\def\todo#1{{\color{red}[TODO: #1]}}
\def\added#1{{\color{magenta}[Added or edited: #1]}}

@@

Causal inference \cite{pearl2009causality} applied to 'omics experiments, elucidates statistical associations \added{(e.g. Pearson correlation, Spearman correlation, mutual information)}, which result from the underlying biomolecular mechanistic relationships. Modern high-throughput technologies such as mass spectrometry-based proteomics quantify components of the biomolecular systems on a large scale.  \added{Thus, the opportunity exists for the proteomics community to apply computational methods for causal inference developed for gene microarray data to proteomics data \cite{bensimon2012mass}.}

We will demonstrate the paradoxical result that the large amount of data generated by high-throughput proteomics experiments make the task of causal inference more difficult.  If only two biologically unrelated analytes were measured simultaneously, there is some probability of a spurious association -- an association between the analytes arising purely from random chance -- though the chance of this happening is low and gets lower as sample size increases. The difficulty of high-throughput data stems from the fact that the more analytes that are measured, the more pairwise associations between analytes are generated, resulting in a much greater chance for the appearance of spurious associations that are just as strong as associations that represent actual biological relationships.  This obfuscates the true signal, leading to increased false discoveries of putative causal events.  \added{While the gene network inference community has developed approaches to dealing with this issue\cite{bensimon2012mass} , the solutions specific to the experimental conditionals and biological questions relevant to proteomics are still in early stages.} The problem can be addressed by refining the biological question, and by improving experimental design in terms of selection of  (1) the subset of analytes, (2) the number of biological replicates, and (3) the type of biological conditions and stresses. Below, we describe in non-technical terms the process of elucidating causal associations from high-throughput data and suggest practical approaches for analysis of large scale \added{proteomics} datasets.

@@

To describe Step 2 of causal inference, we introduce some terminology. In language of \added{probability} the quantified proteins are called {\it variables}. \added{When two proteins are connected in how they vary, such that knowing the behavior of one protein provides information on the other, they are called {\it dependent}.  If there is no such connection, they are called {\it independent}. {\it Conditional independence} is a special case that is important to causal inference.  Two dependent variables are conditionally independent if, after knowing, or {\it conditioning on} the behavior of third-party variables, they become independent -- knowing one provides no additional information about the other after the information from the third party variables is considered.}

\added{Causation and dependence are linked as follows;
\begin{itemize}
\item For two variables, if one is the direct cause of the other, they are dependent, even after conditioning on other variables in the system.
\item If the two variables that are not directly causal related, {\it more often than not are still dependent}, but that dependence disappears after conditioning on the other variables in the system.
\end{itemize}
Note {\it dependence} is a property of the underlying biology, and {\it statistical association} is a property of the data generated by the biology, they are not the same.  {\it Under the right measurement conditions} dependence manifests itself as statistical association in the data and application of statistical analysis can reveal cases of conditional independence (when those conditions breakdown is beyond the scope of this perspective, but Sachs and Itani et al. 2013 \cite{sachs2013single} explores this in detail within the phospho-proteomics context).  So if we had an oracle that revealed that two proteins are still dependent after conditioning on all others, then we'd know they were one of them was a direct causal of the other.  Without an oracle, we must rely on statistical inference of cases where pairs of proteins were conditional independent, ignore those cases and focus on the remaining pairs as potential causal relations.  For a more rigorous treatment of the link between mechanism, probabilistic dependence, and statistics, consult Pearl's seminal work in this area\cite{pearl2009causality}.}

Let's see how this applies to the MAPK signaling cascade. \added{We know from the biology that Raf regulates the abundance of Erk though Mek.  Thus Raf and Erk are dependent.  However Erk abundance is directly determined by Mek, Raf does not exert any additional regulatory influence on Erk beyond its influence {\it through Mek}.  Thus Raf and Erk are conditionally independent given Mek.  This biology is manifests itself in the data.}  Figure \ref{mapkInference}C compares Raf to Erk, indicating in \added{solid circles} the biological samples where concentrations of Mek are high (here, set to the top quartile).   Note that when we subset the data to only the samples with high Mek, we can no longer visually detect the association between Raf and Erk.  \added{To understand why we'd interpret this as evidence for the presence of conditional independence between Raf and Mek, we can build intuition from a formal statistical test of conditional independence -- the null hypothesis is conditional independence between Raf and Erk given Mek, the quantification of association between Raf and Erk {\it after subsetting Mek} is the test statistic, and we decide in favor of the null hypothesis if the p-value for the test statistic (the probability an association at least as high as the one observed could have emerged spuriously) is too high.  Indeed, the class of causal inference algorithms called constraint-based algorithms repeatedly performs this type of test over multiple pairs of variables \cite{spirtes2000causation}. Given this evidence the edge between Raf and Erk is removed, resulting in the middle graph of Figure \ref{mapk}A.}

At Step 2 in the figure, the direction of the regulation remains unknown. Inference of the direction of the chain of events requires that the experimental design involves external interventions or stresses. Figure \ref{mapkInference}D illustrates the results of Step 3, in the case where an intervention targeted Mek with an inhibitor. The intervention does not affect the concentration of Mek, however it blocks its ability to phosphorylate other proteins.  After this intervention the Raf--Mek relationship is unchanged, while Erk drops to a low level.  From this we can infer that Mek has causal influence on Erk, and since Raf was unaffected by the intervention, that Raf has causal influence on Mek.  With the intervention, we can finally move from the undirected graph in panel A - step 2 to the causal graph in panel A - step 3.

In the general case, computational methods for causal inference follow the workflow  in Figure \ref{mapkInference}A, while scaling it to characterize multiple inter-related variables. Step 1 creates a dense network of pairwise associations. \added{Step 2 identifies cases of conditional independence and removes those edges in those cases to create a much sparser network.}  Finally, Step 3 uses the experimental design, specifically the information regarding the interventions, to evaluate these edges as evidence for potential causal events.  See Koller-Friedman \cite{koller2009probabilistic} for a detailed description of these methods and their theoretical underpinnings. Numerous implementations of these algorithms are available in statistical software, e.g. in the R package {\tt bnlearn} \cite{scutari2009learning}.

@@

Similarly, the increased incidence of spurious correlations impedes the performance of statistical methods in Step 2, which elucidate conditional independences in the data.  If we think of this step as \added{attempting to detect which pairs are not conditionally independent}, then the spurious correlations result in more false positives \added{detections}. To illustrate, we repeated the previous simulation, again starting with 20 proteins and 100 biological samples, but this time expanding to only 100 proteins.  Instead of finding the highest spurious correlation between pairs of proteins, we apply an example of type of causal inference algorithm that performs a series of conditional independence tests between the sets of proteins (constraint-based algorithms\cite{spirtes2000causation}), and count the number of pairs proteins \added{where the null hypothesis of conditional independence was rejected}.  As before, since we randomly draw protein concentration measurement values from a Gaussian distribution, the values are completely independent, \added{no test should ever be rejected, and any instance of rejection by this algorithm would have biological justification}.  We again repeat these experiments 500 times. Figure \ref{spur_dep} shows the histograms of the counts of false positives.  The figure graphs the results of the Grow-Shrink algorithm \cite{margaritis2003learning}, \added{but we also repeated the simulation with the IAMB algorithm\cite{tsamardinos2003algorithms} and Inter-IAMB algorithm\cite{yaramakala2005speculative} and achieved similar results}.    The results demonstrate that an increase in the number of proteins leads to an increase in false positives. This means that the computational methods for causal inference will fail for a typical large-scale experiment, because they cannot reliably distinguish the potential causal relationships from noise.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=1\textwidth]{figs/spurious_dep.png}}
\caption{As in Figure \ref{spur_corr}, but reporting the counts of false positives (rejecting the null hypothesis of conditional independence.}
\label{spur_dep}
\end{figure}

A related problem with high-throughput experiments in Step 2 is a relatively small number of biological replicates as compared to the number of analytes. While Step 1 (evaluating the set of pairwise associations) can be carried out with even a small amount of replicates, Step 2 relies on having many replicates relative to the number analytes.  Note the experiments demonstrated in Figure 3 had 100 replicates to its 100 proteins, and still false positives were high.  The quality of causal inference approaches decreases drastically as the numbers of replicates fall.

Finally, Step 3 of the workflow requires interventions to infer causal relationships from evidence of conditional independence.  Quantifying a large number of analytes introduces a  challenge for this step as well. As the number of features grows, the number of interventions needed to fully infer causality grows, and eventually performing a sufficient perturbation experiments becomes infeasible.   We showed with the Raf-Mek-Erk example that one intervention was sufficient to infer two causal relationships. \added{Given theoretical work on the number of interventions necessary to fully elucidate causality between a set of variations\cite{eberhardt2012number}\cite{hauser2012two}\cite{hyttinen2013experiment}}, a set of interventions on $< k$ variables may be sufficient to infer causal events between $k$ analytes . Unfortunately, both the number of biological replicates and the number of interventions im high-throughput profiling experiments is typically small.  Also, if the step 2 results in too many false positives for potential causal relations, this will adversely affect the results of step 3 regardless of the size of the dataset.

@@

\item \textit{Profile more biological replicates.}  The high-throughput measurements should provide more samples from distinct biological sources, which come from a same underlying population, in order to achieve a sufficient statistical power, and distinguish true and spurious associations.  This fact gives advantage to technologies, such as targeted proteomics tools, that quantify fewer analytes but have a higher sample throughput.  \added{A review by Twerve and Saez-Rodriguez provides a contrast of technologies by their content and sampling throughput\cite{SaezRodriguez:2012kx}.}

\added{Single cell mass cytometry is a technology deserving of special mention, where many thousands of cells per sample provide ample statistical power, though cell samples still need to be drawn from multiple individuals in order to make inferences about the population from which those individuals were drawn}. See Sachs et al 2005 for an in depth case study of network inference with a single cell dataset. \cite{sachs2005causal}.

\item \textit{Use prior knowledge.} Prior knowledge improves the search for conditional independence and helps to determine causality. The prior knowledge can be in form of known canonical networks, extracted, e.g. from pathway databases such as KEGG. One example of such prior information is the MAPK pathway. The prior information reduces the search space of unknown associations that need to be considered, enables a more effective use of the data, and increases the confidence in newly discovered statistical associations.  \added{Work by Saez-Rodriguez, Lauffenburger, and Sorger\cite{SaezRodriguez:2009hb}, as well as Terfve and Saez-Rodriguez\cite{Terfve:2015kw} use prior network knowledge to build logic models that reflect causal relationships between signaling proteins using proteomics measurements, the former using antibody/bead-based XMAP technology from Luminex, and the latter using LC-MS/MS.} Another example of prior knowledge is contextual information, such as spatial or temporal annotations of the quantitative measurements in the cell. The contextual information can be extracted from the literature or from other complementary (and potentially noisy) datasets. The causal inference algorithms can be extended to weigh evidence of conditional independence, depending on whether the analytes share the same spatial or temporal context.

\item \textit{Select targeted interventions wisely.} Targeted interventions perturb individual components of the biological system.  An example is \added{siRNA knockdowns, as well as} small molecule inhibitors, which block the causal influence of a specific protein on its downstream components. Although effective, such targeted interventions are limited in number. Therefore, a strategic experimental design would use prior information, prioritize the interventions and the targets, and apply them to parts of the biological system that have most potential for new discovery of regulatory events.  For example, a graph with undirected edges can be inspected, to reveal which nodes have potential to reveal the most causality if perturbed. Such targeted perturbations can be applied iteratively, after an initial statistical analysis revealed areas of the network where causal inference would benefit from extra measurements and data.

\item \textit{Consider broad-scale interventions}. Broad-scale interventions sacrifice specificity of targeted interventions to simultaneously perturb many variables in a biological system. One example of broad-scale interventions is varying experimental conditions, in order to activate multiple pathways.  Signals from endocrine, paracrine, and autocrine ligands elicit various signaling responses in hepatocytes, thus interventions that cover this range of signals gives the best picture of the broader causal network of hepatocyte signaling \cite{alexopoulos2010networks}. Similarly, interventions that go beyond receptor-level and perturb multiple components of the system bring cascading causal direct orientation deeper into the network.  Although they do not provide specific information about the downstream effects of stimulation,  broad-scale interventions can provide more causal insight. Therefore, the advantage of this approach is that it may enable elucidation of  causality across the entire system.
